Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package

### 2020-06-16 050-some-applications.Rmd
update case study with new syntax

```{r processBRCA, echo=TRUE, eval=TRUE, cache=TRUE}
## full workflow simulating RNA-seq data
allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
lastClinical <- which( names(allDat) == 'tumorsize' )
brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
## remove naively the RSEM adjustment
brca <- round(brca, 0)
ncol(brca) ## num of genes 20501
## compute the median expression! avoid the 0s
brcaMedian <- apply(brca, 2, median)
## retain top (1-probs)*100% highest expressing genes for illustration
myProb <- 0.95 
cutPoint <- quantile( x = brcaMedian, probs = myProb )
## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
brca <- brca[ , genesToKeep ]
## ncol(brca) / 20501
(d <- length(genesToKeep))
```

@@

```{r estMargins, echo = TRUE, eval = TRUE, cache=TRUE}
## 2. Estimate NegBin parameters using Method of Moments
estimateNegBinMoM <- function(tmpGene, minP = 1e-5) {
    tmpMean <- mean(tmpGene)
    tmpVar <- var(tmpGene)
    ## relate to nbinom parameters
    ## See ?rbinom for details.
    p <- tmpMean / tmpVar
    if (p < minP) {p <- minP } ## maybe add noise here
    n <- ( tmpMean * p) / ( 1  - p )  
    ## format for bigsimr margins
    return( list("nbinom", size = n, prob = p) )
}
brcaMargins <- apply( unname(as.matrix(brca)), 2, estimateNegBinMoM )
## describe the margins
## check for too small prob
## head( sort( unlist( lapply( brcaMargins, function(x) {x$prob} ) ) ) )
## head( sort( unlist( lapply( brcaMargins, function(x) {x$size} ) ) ) )
```

```{r simBRCA, echo=TRUE, eval = FALSE, cache=TRUE}
## 3. Generate the simulated samples
## N <- 10
## system.time( simBRCA <- rvec(N, rho = rho[1:2, 1:2], params = brcaMargins[1:2], type = corType, adjustForDiscrete = FALSE) )
N <- nrow(brca)
system.time( simBRCA <- rvec(N, rho = rho, params = brcaMargins, cores = CORES, type = corType, adjustForDiscrete = FALSE) )
simBRCA[1:2, 1:2 ]

>>>>

```{r processBRCA, echo=TRUE, eval=TRUE, cache=TRUE}
## full workflow simulating RNA-seq data
allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
lastClinical <- which( names(allDat) == 'tumorsize' )
brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
## remove naively the RSEM adjustment
brca <- round(brca, 0)
ncol(brca) ## num of genes 20501
## compute the median expression! avoid the 0s
brcaMedian <- apply(brca, 2, median)
## retain top (1-probs)*100% highest expressing genes for illustration
myProb <- 0.99
cutPoint <- quantile( x = brcaMedian, probs = myProb )
## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
brca <- brca[ , genesToKeep ]
## ncol(brca) / 20501
(d <- length(genesToKeep))
```

@@

```{r estMargins, echo = TRUE, eval = TRUE, cache=TRUE}
## 2. Estimate NegBin parameters using Method of Moments
estimateNegBinMoM <- function(tmpGene) {
    tmpMean <- mean(tmpGene)
    tmpVar <- var(tmpGene)
    ## relate to nbinom parameters
    ## See ?rbinom for details.
    p <- tmpMean / tmpVar
    n <- ( tmpMean * p) / ( 1  - p )  
    ## format for bigsimr margins
    get( 'qnbinom')
    return( str2lang( paste0( "qnbinom(size = ", n, ", prob = ", p, ")" ) ) )
}
brcaMargins <- apply( unname(as.matrix(brca)), 2, estimateNegBinMoM )
## brcaMargins <- apply( unname(as.matrix(brca))[ , 1:3], 2, estimateNegBinMoM )
## describe the margins
## check for too small prob
## head( sort( unlist( lapply( brcaMargins, function(x) {x$prob} ) ) ) )
## head( sort( unlist( lapply( brcaMargins, function(x) {x$size} ) ) ) )
```

```{r simBRCA, echo=TRUE, eval = TRUE, cache=TRUE}
## 3. Generate the simulated samples
N <- nrow(brca)
system.time( simBRCA <- rvec(n = N, rho = rho, margins = brcaMargins, cores = CORES, type = corType) )
simBRCA[1:2, 1:2 ]

### 2020-06-16 index.Rmd
update case study with new syntax

--- 
title: "Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package"
author: [A.G. Schissler, A. Knudson]
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
site: bookdown::bookdown_site
bibliography: [bigsimr.bib, packages.bib]
biblio-style: apalike
link-citations: yes
abstract: |
  In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies. This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional. This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing). Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.  Major challenges lie in the computational complexity involved in simulating these massive random vectors. We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics. As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets. The proposed algorithm is implemented as the `bigsimr` R package.
---

>>>>

--- 
title: "Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package"
author: [A.G. Schissler, A. Knudson, T. Kozubowski, A. Panorksa]
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
site: bookdown::bookdown_site
bibliography: [bigsimr.bib, packages.bib]
biblio-style: apalike
link-citations: yes
abstract: |
  In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies. This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional. This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing). Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.  Major challenges lie in the computational complexity involved in simulating these massive random vectors. We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics. As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets. The proposed algorithm is implemented as the `bigsimr` R package.
---
